---
title: "hmw05 - yavyx"
author: "Javier Castillo-Arnemann"
date: "October 19, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we start by loading the necessary packages.

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
library(gapminder)
library(knitr)
```

##Part 1: Factor Management
We will first check the structure of the `gapminder` dataset to check that we have factors in it.
```{r}
str(gapminder)
```

As we can see, there are 2 factors: `country` with 142 levels, and `continent` with 5 levels.

We will now filter Oceania out of the dataset and see if the levels change.
```{r}
gapm_wo_oceania <- gapminder %>%
  filter(continent != "Oceania")
str(gapm_wo_oceania)
```

Even though Oceania' countries are not in the new dataset, it still retains the levels. We can drop these levels with base R, or the `forcats` package.
```{r}
gapm_wo_oceania %>%
  droplevels() %>%
  str()

gapm_wo_oceania %>%
  mutate(continent = fct_drop(continent)) %>%
  mutate(country = fct_drop(country)) %>%
  str()
```
As we can see, we can drop all levels with base R, and with `forcats`, we can drop individual ones. However, the base R function has an `except` argument that lets us specify which factor we don't want to drop. 

```{r}
gapm_wo_oceania %>%
  droplevels(except = 2) %>%  #Indicate column to keep
  str()
```

Here we see that only the empty `country` factors were dropped, beacuse the second factor (continent) was specified in the `except` parameter.

We can also reorder the factor other than alfabetically (which is the default). We'll work with American countries for this.

When we plot some information about the countries, they are ordered alphabetically.
```{r}
gap_america <- gapminder%>%
  filter(continent == "Americas") %>%
  droplevels()

ggplot(gap_america) +
    geom_bar(aes(country, lifeExp/n_distinct(gapminder$year)), stat = "identity", fill = "steel blue") + #tell geom_bar to use the y aesthetic instead of the bar. Divide by 12 to get correct lifeExp values
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #change x label angles to be more readable
    labs(title = "Life Expectancy in the Americas",
         x = "Country", y = "Life Expectancy")
```

We can change this by reordering the countries factor by the population in each country, for example
```{r}
gap_america %>%
  mutate(country = fct_reorder(country, pop)) %>% #Reorder factor
  ggplot() +
    geom_bar(aes(country, lifeExp/n_distinct(gapminder$year)), stat = "identity", fill = "steel blue") + #tell geom_bar to use the y aesthetic instead of the bar. Divide by 12 to get correct lifeExp values
    theme(axis.text.x = element_text(angle = 60, hjust = 1)) + #change x label angles to be more readable
    labs(title = "Life Expectancy in the Americas (ordered by population)",
         x = "Country", y = "Life Expectancy")
```

We can see that in general, there is not a strong relationship between population and life expectancies in the Americas.

Let's see if by using `arrange()` we get similar results.

```{r}
gap_america_arr <- gap_america %>%
  droplevels() %>%
  group_by(country) %>%
  mutate(mean_pop = mean(pop)) %>%
  arrange(mean_pop) #Arrange by mean poplutation some all country measurements stay

kable(head(gap_america_arr, 15))

levels(gap_america_arr$country)
```

Here we see that the dataset is arranged by population too, but with the `levels` function we can see that the factor is still ordered alfabetically, so when we plot it we will see the same.
```{r}
ggplot(gap_america_arr) +
    geom_bar(aes(country, lifeExp/n_distinct(gapminder$year)), stat = "identity", fill = "steel blue") + #tell geom_bar to use the y aesthetic instead of the bar. Divide by 12 to get correct lifeExp values
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #change x label angles to be more readable
    labs(title = "Life Expectancy in the Americas",
         x = "Country", y = "Life Expectancy")
```

Looking in the [R documentation](https://www.rdocumentation.org/packages/dplyr/versions/0.7.6/topics/arrange), we can see that `arrange()` only reorders the rows of a dataframe, so it will never change the factor levels order which is what we are looking for.

##Part 2: File I/O
We'll do some experimenting with reading in data, modifying it, and seeing if the changes remain after saving it to a new file and reading it in again.

First, let's filter the gapminder dataset to only have information from 2007 and save that in a csv file, and read it back in to see if it keeps the factors' order.
```{r}
gapminder_2007 <- gapminder %>%
  filter(year == 2007)
str(gapminder_2007)

write_csv(gapminder_2007, "gapminder2007.csv") 

gapminder_2007_2 <- read_csv("gapminder2007.csv")
str(gapminder_2007_2)
```

We can see that `read_csv()` does not coerce character values into factores, like the base R `read.csv` function does. Therefore, we lose the factors' information and would have to manually convert the variables factors as follows.
```{r}
gapminder_2007_2$country <- factor(gapminder_2007_2$country)
gapminder_2007_2$continent<- factor(gapminder_2007_2$continent)

str(gapminder_2007_2)
```

Let's check if we can keep the `arrange()` does survive the writing and reading steps.

```{r}
gapminder_2007_arr <- gapminder_2007 %>%
  arrange(continent)

head(gapminder_2007_arr)

write_csv(gapminder_2007_arr, "gapminder2007_arr.csv")

gapminder_2007_arr_2 <- read_csv("gapminder2007_arr.csv")

head(gapminder_2007_arr_2)
```

This change does survive the writing/reading steps.

Finally, we can check if after reordering a factor we would get the same ordering with `read.csv()` that does coerce character vectors into factors.
```{r}
gapminder_2007 <- gapminder_2007 %>%
  mutate(country = fct_reorder(country, pop))

str(gapminder_2007)

write_csv(gapminder_2007, "gapminder2007_factors.csv") 

gapminder_2007_2 <- read.csv("gapminder2007_factors.csv")

str(gapminder_2007_2)

```

Our new dataset does have the factors as we predicted, but the order was not kept. This is probably because `.csv` files don't keep the factor information, the are just plain text files and there is no way that R can recover this information. R assigns the factors after reading in the data.

##Part 3: Visualization design
We'll modify one of the life expectancy figures from part 1.

First attempt:
```{r}
ggplot(gap_america) +
    geom_bar(aes(country, lifeExp/n_distinct(gapminder$year)), stat = "identity", fill = "steel blue") + #tell geom_bar to use the y aesthetic instead of the bar. Divide by 12 to get correct lifeExp values
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + #change x label angles to be more readable
    labs(title = "Life Expectancy in the Americas",
         x = "Country", y = "Life Expectancy")
```

New version:
```{r}
final_plot <- ggplot(gap_america) +
    geom_bar(aes(country, lifeExp/n_distinct(gapminder$year), fill = lifeExp > mean(lifeExp)), stat = "identity") + #tell geom_bar to use the y aesthetic instead of the bar. Divide by 12 to get correct lifeExp values. Color bars according to times lifeExp was above the mean
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid.major = element_blank()) + #change x label angles to be more readable
    labs(title = "Life Expectancy in the Americas",
         x = "Country", y = "Life Expectancy") +
    scale_y_continuous(breaks = 1:10 * 10) + #add y labels every 10 years
    scale_fill_discrete(name = "Life Expectancy",
                         breaks = c("FALSE", "TRUE"),
                         labels = c("< 65.7", "> 65.7")) #edit legend values

final_plot
```

Finally, we'll nake a 3D plotly that shows similar information as the previous plot, and save it.
```{r}
plot_ly(gap_america, 
        x = ~country, 
        y = ~lifeExp, 
        z = ~year,
        type = "scatter3d",
        mode = "markers",
        opacity = 0.8,
        color = ~ lifeExp > mean(lifeExp)) %>%
  htmlwidgets::saveWidget("plotly.html")

```

##Part 4: Writing figures to file
We will save the last bar plot to a file using `ggsave`.

```{r}
ggsave("lifeExp_america.png", plot = final_plot, width = 18, height = 8, units = "cm")
```


